# Event objects {#event-chapter}

```{r,echo=FALSE,message=FALSE}
source("setup.R")
```

Event objects are similar to the data sets described in \@ref(data-set-chapter), 
but are simpler and easier to create. This is the fastest way to implement a 
basic intervention (like dosing) for a single "individual" into your model. 

Event objects also offer an elegant way to compose complicated dosing regimens. 
Typically, the different parts of a regimen are composed as individual event 
objects and then combined to create a multi-faceted dose regiment. 

Finally, once an event object is created (either simple or complex), this 
object can be "expanded" into multiple individuals to create a population 
data set for simulation. 

See details in the subsequent sections.

## Usage 

Event objects are frequently used in a pipeline to simulate a dosing regimen. 
For example

```{r}
mod <- house(end = 72) 

mod %>% ev(amt = 100, ii = 24, addl = 1) %>% mrgsim() %>% plot("CP")
```

This used the `ev()` constructor to  make an event object for two 100 mg doses 
and this is passed into `mrgsim()` to implement this regimen. 

Alternatively, we can create a standalone object and feed that into the pipeline

```{r, eval = FALSE}
regimen <- ev(amt = 100, ii = 24, addl = 1)

mod %>% ev(regimen) %>% mrgsim() %>% plot("CP")
```

If you are not using the pipe syntax, the following would be equivalent 
calls

```{r, eval = FALSE}
mrgsim(mod, events = regimen) %>% plot("CP")
```

And there are `mrgsim()` variants that explicitly accept an event object

```{r, eval = FALSE}
mrgsim_e(mod, regimen) %>% plot("CP")
```

More will be said about how to create and manipulate event objects in the 
following sections. 




## Construction

A new event object can be created with the `ev()` constructor. For a single, 
100 mg dose it would be

```{r}
e <- ev(amt = 100)
```

When you print the object to the R console we see the 100 mg dose along with 
the following defaults

- `time` set to 0
- `cmt` set to 1 (the first compartment)
- `evid` set to 1 (a bolus dose)

```{r}
e
```

Of course, we can override any of these defaults or add additional items as
needed. For a single 100 mg dose infused over 2 hours in compartment 2 one 
hour after the simulation starts

```{r}
e <- ev(amt = 100, rate = 50, cmt = 2, time = 1)
```

To use this event object, we can pass it into `mrgsim()` under the `events` 
argument

```{r}
mod <- house(delta = 1, end = 24)

mrgsim(mod, events = e)
```

Event object inputs can be functions of previously defined inputs. For example

```{r}
ev(amt = 100, rate = amt / 2)
```


See the `?ev()` help topic for more information on additional arguments 
when constructing event objects. Here, I'd like to specifically highlight a 
handful of options that can be helpful when constructing event objects. 

__Infusion duration__

Above, we created some infusion event objects by adding an infusion rate to the 
input. We can also indicate an infusion by adding an infusion time through 
the `tinf` argument

```{r}
ev(amt = 100, tinf = 2)
```


__ID__

While the primary use case for event objects are for _single individuals_, 
we can code a series of IDs into the object too

```{r}
ev(amt = 100, ID = 1:3)
```

Here, we asked for 3 IDs in the object. Once this is turned into a simulation
data set (see below), we'll have a population data set from which to simulate.


__Additional data items__

We can also pass through arbitrary data columns through the event object. For
example, we can pass through `WT`

```{r}
ev(amt = 100, WT = 80)
```

## Coerce to data set

As we noted, event objects are very similar to data sets and they are nothing 
but data sets under the hood. We can take the event objects we created above
and coerce them to other objects. 

__Using as_data_set__

```{r}
as_data_set(e)
```


This will ensure that there is an `ID` column in the output and it will 
be suitable to use for simulation.

__Using as.data.frame__

```{r}
as.data.frame(e) %>% mutate(ID = 5)
```

## Extract information

There is a `$` operator for event objects

```{r}
e$amt
```


## Combining event objects


### Concatenate

Two or more event objects can be concatenated using the `c` operator

```{r}
e1 <- ev(amt = 100)
e2 <- ev(amt = 200, time = 24)

c(e1, e2)
```

This essentially "rbinds" the rows of the individual event objects and 
sorts the rows by `time`. 

__NOTE__: the result of this manipulation is another event object. 

### Sequence

Event objects can also be combined to happen in a sequence. In the previous
example, we wanted the 200 mg to happen at 24 hours and we had to code that
fact into `time` accordingly.

By specifying a dosing interval (`ii`) we can ask mrgsolve to do that 
automatically by calling the `seq()` method.

```{r}
e1 <- ev(amt = 100, ii = 24)
e2 <- ev(amt = 200, ii = 24)

seq(e1, e2)
```

This was a trivial example to get a simple result. We can try something more 
complicated to make the point

```{r}
e3 <- ev(amt = 100, ii = 6,  addl = 28)
e4 <- ev(amt = 200, ii = 12, addl = 124)
e5 <- ev(amt = 400, ii = 24, addl = 3)

seq(e3, e4, e5)
```

__NOTE__: when mrgsolve puts event objects into a sequence, it starts the 
next segment of the regimen one dosing interval after the previous regimen 
finished. Going back to the simple example

```{r}
seq(e1, e2)
```

`e1` was just a single dose at time 0. mrgsolve will have `e2` start one dosing
interval (24 hours) after the last (only) dose in `e1`. We can alter the amount
of time between segments of the regimen by using the `wait` argument. For 
example, to push `e2` out by an additional 24 hours we'd use

```{r}
seq(e1, wait = 24, e2)
```

We can also use a negative value for `wait` to make the next dose happen sooner

```{r}
seq(e1, wait = -12, e2)
```


Finally, we should note that event objects can be used multiple times in a 
sequence

```{r}
seq(e1, e2, wait = 7*24, e2, e1)
```

### repeat

Like the `seq()` method for event objects, `ev_repeat` will put an event object
into a sequence `n` times

```{r}
ev_repeat(e1, n = 3)
```

By default, this function returns a regular data frame. To return an event 
object instead call

```{r, eval = FALSE}
ev_repeat(e1, n = 3, as.ev = TRUE)
```


You can put a waiting period too. To illustrate this, let's compose a more 
complicated regimen and repeat that

```{r}
e1 <- ev(amt = 500, ii = 24)
e2 <- ev(amt = 250, ii = 24, addl = 5)
e3 <- ev_seq(e1, e2)

e3 %>% realize_addl()
```

In this regimen, we have daily dosing for 7 doses. The last dose is given 
at 144 hours. When putting this into a sequence, we'll wait one dosing interval
and then the `wait` period and then start again

```{r}
ev_repeat(e3, n = 3, wait = 7*24)
```


### Create a "data_set"

Use the `as_data_set()` function to combine multiple event objects into a single
data set. 

```{r}
as_data_set(e1, e2)
```

It's important to note that 

1. The result is a regular old `data.frame()`; once you call `as_data_set()`, 
   you exit the event object world
1. Each event object is given a different `ID`


Recall that we can create event objects with multiple `IDs`; `as_data_set()`
is handy to use with this feature

```{r}
as_data_set(
  ev(amt = 100, ID = 1:3), 
  ev(amt = 200, ID = 1:3), 
  ev(amt = 300, ID = 1:2)
)
```


Notice that `as_data_set` has created unique IDs for the 3 subjects in 
the 100 mg group, the 3 subjects in the 200 mg group, and the 2 subjects in 
the 300 mg group.

We'll cover a function called `ev_rep()` below to "expand" an event object to 
multiple individuals

```{r, eval = FALSE}
as_data_set(
  e1 %>% ev_rep(1:300),
  e2 %>% ev_rep(1:300)
)
```


## Modifying an event object

### Tidy-like manipulation

Event objects can be mutated

```{r}
mutate(e, amt = 200)
```

Columns can be removed from event objects

```{r}
ev(amt = 100, WT = 50, AGE = 12) %>% select(-WT)
```

Rows can be removed from event objects

```{r}
e <- c(ev(amt = 100), ev(amt = 200, time = 12), ev(amt = 300, time = 24))

filter(e, time <= 12)
```


### realize_addl

"Additional" doses can be made explicit in an event object

```{r}
ev(amt = 100, ii = 6, addl = 3) %>% realize_addl()
```

### ev_rep

Event objects can be "expanded" into multiple IDs to create a population; 
use the `ev_rep()` function for this. 

```{r}
ev(amt = 100) %>% ev_rep(1:5)
```

By default, `ev_rep()` returns a regular data frame. You can request that 
an event object is returned

```{r, eval = FALSE}
ev(amt = 100) %>% ev_rep(1:5, as.ev = TRUE)
```

`ev_rep()` can work on an event object with any complexity.


## Creative composition

mrgsolve has a couple of more creative ways to construct event objects. 

### ev_days

`ev_days()` will create dosing sequences when dosing are on certain days (of the 
week). For example, to dose only on Monday, Wednesday, and Friday for on month

```{r}
e <- ev_days(ev(amt = 100), ii = 168, addl = 3, days = 'm,w,f')
e
```

We can see how this works by simulating the regimen

```{r}
mrgsim_e(mod, e, end = 168*4) %>% plot("CP")
```


### ev_rx

`ev_rx()` is a way to write a regimen out with notation similar to what you 
might see on a prescription. For example, 100 mg twice daily for 3 doses into 
compartment 2 would be

```{r}
ev_rx("100 mg q12h x3 in 2")
```

To code an infusion

```{r}
ev_rx("500 mg over 2 hours q 24 h x3 in 1")
```


See the `ev_rx()` documentation for more details and limitations.

