# Event objects {#event-chapter}

```{r,echo=FALSE,message=FALSE}
source("setup.R")
```

Event objects are similar to the data sets described in \@ref(data-set-chapter), 
but are simpler and easier to create. This is the fastest way to implement a 
basic intervention (like dosing) into your model. 

## Construction

A new event object can be created with the `ev()` constructor. For a single, 
100 mg dose it would be

```{r}
e <- ev(amt = 100)
```

When you print the object to the R console we see the 100 mg dose along with 
the following defaults

- `time` set to 0
- `cmt` set to 1 (the first compartment)
- `evid` set to 1 (a bolus dose)

```{r}
e
```

Of course, we can override any of these defaults or add additional items as
needed. For a single 100 mg dose infused over 2 hours in compartment 2 one 
hour after the simulation starts

```{r}
e <- ev(amt = 100, rate = 50, cmt = 2, time = 1)
```

To use this event object, we can pass it into `mrgsim()` under the `events` 
argument

```{r}
mod <- house(delta = 1, end = 24)

mrgsim(mod, events = e)
```

Event object inputs can be functions of previously defined inputs. For example

```{r}
ev(amt = 100, rate = amt / 2)
```


See the `?ev()` help topic for more information on additional arguments 
when constructing event objects. Here, I'd like to specifically highlight a 
handful of options that can be helpful when constructing event objects. 

__Infusion duration__

Above, we created some infusion event objects by adding an infusion rate to the 
input. We can also indicate an infusion by adding an infusion time through 
the `tinf` argument

```{r}
ev(amt = 100, tinf = 2)
```

__Additional data items__

We can also pass through arbitrary data columns through the event object. For
example, we can pass through `WT`

```{r}
ev(amt = 100, WT = 80)
```

## Coerce to data set

As we noted, event objects are very similar to data sets and they are nothing 
but data sets under the hood. We can take the event objects we created above
and coerce them to other objects. 

__Using as_data_set__

```{r}
as_data_set(e)
```


This will ensure that there is an `ID` column in the output and it will 
be suitable to use for simulation.

__Using as.data.frame__

```{r}
as.data.frame(e) %>% mutate(ID = 5)
```

## Extract information

There is a `$` operator for event objects

```{r}
e$amt
```


## Combining event objects

Coming soon.

## Modifying an event object

Event objects can be mutated

```{r}
mutate(e, amt = 200)
```

Columns can be removed from event objects

```{r}
ev(amt = 100, WT = 50, AGE = 12) %>% select(-WT)
```

