```{r, echo = FALSE, message = FALSE, include = FALSE}
source("setup.R")
```

# Internals {#internals}

This chapter describes some of the mrgsolve internals, including the model 
object, the model itself, and the workflow for building the model. This chapter
is a work in progress that will continue to be developed as time allows. 

__NOTE__ Some of the code chunks in the following sections will be evaluated, 
and some won't. If you don't see output, that means it was suppressed but you 
can try the code on your own.

## Model object

The model object is an S4 object. The object has the following slots


```{r}
mod <- modlib("pk1")

slotNames(mod)
```

### Slots

__model__

Type: `<character>`

This is the name of your model

```{r}
mod@model
```

The build location and several internal C++ functions are named according to 
this model name; mrgsolve will assume that two models with the same name are 
just different iterations of the same model.

__package__

Type: `<character>`

This is the name of the compiled shared object. 

```{r}
mod@package
```

Note that the `house()` model is compiled into the mrgsolve package shared 
object, so `package` for that model is mrgsolve

```{r}
house()@package
```

__soloc__

Type: `<character>`

This is the location of the shared object; it is where the model is actually 
compiled. 

```{r}
mod@soloc
```

You can change the model build location with the `soloc` argument to `mread()`.

__project__

Type: `<character>`

This is the location of the model source files. We are using an example model
from an internal library, so the location will be where the mrgsolve package
is installed. 

```{r}
mod@project
```

For most applications, this will be the location where you create and save
your model file(s).

__code__

Type: `<character>`

This is the source code from the model file

```{r}
head(mod@code)
```

__start__, __end__, __delta__, __add__

Type: `<numeric>`

These form the internal simulation time grid

```{r}
mod@start
mod@end
mod@delta
mod@add
```

__param__

This is an object of class `parameterlist`
```{r}
class(mod@param)
mod@param
```

The `parameterlist` inherits from `numericlist`

```{r}
inherits(mod@param, "numericlist")
```

The parameter list can be extracted with param

```{r, eval = FALSE}
param(mod)
```

The parameter list can be coerced to an R list

```{r, eval = FALSE}
as.list(mod@param)
```

