# Simulated output {#simulated-output}

## Output types

When `mrgsim()` is used to simulate from a model, it by default returns an object with class 
`mrgsims`.  This is an S4 object containing a data.frame of simulated output and a 
handful of other pieces of data related to the simulation run that can be coerced to other
types (like `data.frame` or `tibble`). 

For simulations with large outputs or extremely brief simulations where efficiency is 
important, users can request the output be returned as a data frame. This is most efficient
when the features provided by the `mrgsims` object are not needed. To do this, pass
the `output` argument to `mrgsim()`

```{r, eval = FALSE}
out <- mrgsim(mod, ..., output = "df")
```
or use `mrgsim_df()`

```{r, eval = FALSE}
out <- mrgsim_df(mod, ....)
```

## Methods for mrgsim output

`mrgsolve` provides several methods for working with `mrgsims` objects or 
coercing the simulation matrix into other `R` objects. Note the discussion in the 
following subsections all refer to working with `mrgsims` objects, not data.frame output.

### Coercion methods

* `as_tibble()`: convert to `tibble`
* `as.data.frame()`: convert to `data.frame`
* `as.matrix()`: convert to `matrix`

### Query methods
* `head()`: shows the first `n = 5` rows
* `tail()`: shows the last `n = 5` rows
* `names()`: shows the column names
* `dim()`: shows the number of rows and columns
* `summary()`: shows a numeric summary of all columns
* `$`: extracts a column

### Graphical methods

There is a `plot()` methods for simulated output that is aware of independent and dependent
variables from the simulation. If `out` is the simulated output (an `mrgsims` object)

```{r, eval = FALSE}
plot(out)
```
Plot with a formula; the following example selects only the `CP` and `RESPONSE` outputs
and plots them versus time

```{r, eval = FALSE}
plot(out, CP + RESPONSE ~ time)
```

To select a large number of responses to plot, pass a character vector or comma-separated
character data containing output columns to plot

```{r, eval = FALSE}
plot(out, "CP, RESPONSE, WT, DOSE")
```


### Methods for `dplyr` verbs
`mrgsolve` provides several S3 methods to make it possible to include `dplyr` verbs in 
your simulation pipeline. 

For example

```{r, eval = FALSE}
library(dplyr)
library(mrgsolve)

mod <- mrgsolve:::house()

mod %>% 
  ev(amt=100) %>%
  mrgsim() %>% 
  filter(time >= 10)
```

Here, `mrgsim()` returns an `mrgsims` object.  When `dplyr` is also loaded, this object 
can be piped directly to `dplyr::filter()` or `dplyr::mutate()` etc.

It is important to note that when `mrgsims` output is piped to `dplyr` functionality, 
it is coerced to `tibble` (`data.frame`) and there is no way to get the data back to 
`mrgsims` object. Most of the time, this is desirable and there is no need to explicitly
coerce to `tibble()` when calling `dplyr` verbs on simulated output. 


Other `dplyr` functions that can be used with `mrgsims` objects

* `group_by()`
* `mutate()`
* `filter()`
* `summarise()`
* `select()`
* `slice()`
* `pull()`
* `distinct()`
* `slice()`

### Modify methods

You can modify the underlying data in the `mrgsims` object and keep it as an `mrgsims` object.

- `filter_sims()`: calls `dplyr::filter()` to pick rows to keep or discard
- `select_sims()`: calls `dplyr::select()`; note that `ID` and `time` columns are always retained
- `mutate_sims()`: calls `dplyr::mutate()` to add or modify columns

