```{r,echo=FALSE,message=FALSE}
source("setup.R")
```

# Questions and Answers {#q-and-a}

I'm using this chapter as a place to provide miscellaneous information that 
might not have an obvious place to live. I'd call this a FAQ, but not all of 
the questions are asked frequently or at all. 

## Can I interrupt a simulation?

Starting with `mrgsolve` version `0.11.1`, you can interrupt a long simulation
by pressing `Control-C` or `Esc`, the standard way to pass an interrupt signal
through R. `mrgsolve` will stop every so often to look for the interrupt 
signal.  

You can control the frequency with which `mrgsolve` looks for the interrupt 
signal through an argument to `mrgsim` (default: 256 simulation records).
Increase to check less frequently, increase to check more frequently (this might
be needed for a model where a large amount of work is required to advance
one step) or set to negative number to never check.

## Can I pass compiler flags to my model?

Compiler flags can be passed to your model by setting  `PKG_CXXFLAGS` in 
`$ENV`.  For example

```{c, eval = FALSE}
$ENV
PKG_CXXFLAGS = "-std=c++11"
```

will compile your model according to `C++11` standard (but note that there is a
special plugin that will do this automatically for you; see Section 
\@ref(plugin-cxx11)).

## Can I compile my model with `C++11`?

Yes, you can do this by invoking the `CXX11` plugin (Section \@ref(plugin-cxx11)).

## How can I calculate time after dose?

There are three approaches

__tad argument to mrgsim()__

To get time after dose into your output you can call

```{r, eval = FALSE}
mrgsim(mod, tad = TRUE)
```

and the output will have a `tad` column. Note this does not let you interact
with the `tad` value inside your model.

__Simple calculation in the model__

Most applications will call `self.tad()` (Section \@ref(self.tad)). For example

```{c, eval = FALSE}
[ main ] 
double tad = self.tad();
```

__More complicated calculation in the model__

You can get more control and track `tad` in a specific compartment by using 
the `tad` plugin.  See Section \@ref(plugin-tad) for details.

## My model failed to compile; what now?

The model can fail to compile for a variety of reasons, including an error 
in the `C++` code or inability of R to find the compiler and other pieces of
the tool chain. 

If your model is not compiling, try the `recover` argument to `mread()` (or 
`mcode()`)

```{r, eval = FALSE}
mod <- mread(..., recover = TRUE)
```

You will see a warning on the console and `mread()` will return a list of 
build information. You can look into that information or share it in the 
[mrgsolve issue tracker](https://github.com/metrumresearchgroup/mrgsolve/issues) 
on GitHub.

If your model has `C++` syntax problems, the errors should be printed on the 
console. If you possibly have problems with the compiler or the rest of the 
toolchain, take a look at the `pkgbuild` package, which provides some helpful
tools, especially if you are working on a Windows platform

```{r, eval = FALSE}
pkgbuild::check_build_tools()
pkgbuild::has_build_tools()
pkgbuild::has_rtools()
```


## Can I run mrgsolve on a network drive?

No; do not run mrgsolve on a network drive. Your R installation, mrgsolve 
installation, and R working directory should be on a local hard disk.

## Can I run mrgsolve on a cloud-synced folder?

No; do not run mrgsolve in a synced folder for cloud services like OneDrive, 
GoogleDrive, DropBox etc. Your R installation, mrgsolve installation, and R 
working directory should be on a local hard disk. 

## Can I run mrgsolve in a path that includes spaces?

No; do not run mrgsolve in a path that includes spaces. Your R installation, 
mrgsolve installation, and R working directory should all be in locations 
whose paths do not include spaces.