```{r,echo=FALSE,message=FALSE}
source("setup.R")
```

# Model specification

This chapter details the `mrgsolve` model specification format.  

## How / where to write a model
There are two ways to write your model:

### Separate file {#spec-separate-file}
Open a text editor and type the model into a file with name that has the 
format `<model-name>.cpp`.  This filename format identifies a "name" for your 
model (`<model-name>`, the "stem" of the file name).  The extension __MUST__ 
be `.cpp` (`mrgsolve` curently assumes the extension).  Note: this whole file 
will be read and parsed, so everything in it must be valid `mrgsolve` model
specification elements.  

Use the `mread()` function to read and parse this file.  For the model 
called `mymodel` saved in `mymodel.cpp` (in the current working directory), 
issue the command:
```{r,eval=FALSE}
mod <- mread("mymodel")
```

`mread()` returns a model object from which you can simulate.



### Inline / `code` \ {#spec-inline}
Often it is more convenient to write a model right in your `R` script.  The 
model might look something like this:

```{r}
code <- '
$PARAM CL = 1, VC = 20
$PKMODEL ncmt=1
'
```

Here, we created a character vector of length 1 and saved it to the `R` object
called `code`.  The name of this object is irrelevant.  But `code` will be 
passed into `mrgsolve` as the model definition.  When `mrgsolve` gets a model 
like this along with a "name" for the model, mrgsolve will write the code to a 
file called `<model-name>.cpp` and read it right back in as if you had typed 
the code into this file (section \@ref(spec-separate-file)). 

To parse and load this model, use the `mcode()` command:

```{r,eval=FALSE}
mod <- mcode("mymodel", code)
```

`mcode()` is a convenience wrapper for `mread()`.  `mcode` writes the code to
`mymodel.cpp` in `tempdir()`, reads it back in, compiles and loads.

The `mcode` call is equivalent to:
```{r, eval=FALSE}
mod <- mread("mymodel", tempdir(), code)
```

For help, see `?mread` , `?mcode` in the `R` help system after loading 
`mrgsolve`.



## Code blocks {#code-blocks}

### About code blocks

__Block identifier__
The `mrgsolve` model specification involves several different types of code 
that are written in different blocks.  A block starts with an identifier
that starts with `$BLOCKID`, where `BLOCKID` is a character designator.  For 
example, parameters are written in `$PARAM`, differential equations are written
in `$ODE` etc.  Users are free to include block code on the 
same line as the block identifier, but must include a space after the identifier. 
For example, the parser will recognize `$PARAM CL = 1` but not `$PARAMCL=1` as parameters.

Different blocks may require different syntax.  For example, code written in
`$PARAM` will be parsed by the `R` parser and will generally need to adhere 
to `R` syntax requirements.  On the other hand, code in `$MAIN`, `$ODE`, 
and `$TABLE` will be used to make functions in `C++` and therefore will 
need to be valid `C++` code, including terminal `;` on each line.  


__Block options__
Options may be specified on some code blocks that signal how the code is 
to be parsed or used in the simulation.  


### `$PARAM` \ {#block-param}
Define the parameter list in the current model. 

Example:
```{c, eval=FALSE}
$PARAM CL = 1, VC = 20, KA = 1.2
KM = 25, VMAX = 400, FLAG = 1, WT = 80
SEX = 0, N = sqrt(25)
```

Annotated example:
```{c, eval=FALSE}
$PARAM @annotated
CL :   1 : Clearance (L/hr)
VC :  20 : Volume of distribution (L)
KA:  1.2 : Absorption rate constant (1/hr)
```

Notes:

* Multiple blocks are allowed
* Values are evaluated by the `R` interperter


See also: section \@ref(block-theta) and \@ref(block-fixed).

See `?param` in the `R` help system after loading `mrgsolve`.

### `$FIXED`  \ {#block-fixed}
Like `$PARAM`, `$FIXED` is used to specify `name=value` pairs.  Unlike 
`$PARAM`, however, the values associated with names in `$FIXED` are not able 
to be updated.  

By default, names in `$FIXED` are associated with their value through a 
`C++` preprocessor `#define` statement. 

Usually, `$FIXED` is only used when there are a very large number of 
parameters ($>$ 100 or 200).  When some of these parameters never need 
to be updated, you can move them to a `$FIXED` block to get a modest gain 
in efficiency of the simulation.

Items in `$FIXED` will not be shown when parameters are queried. 

Example:
```{c,eval=FALSE}
$PARAM CL = 2, VC = 20

$FIXED
g = 9.8
```


Annotated example:
```{c, eval=FALSE}
$FIXED @annotated
g : 9.8 : Acceleration due to gravity (m/s^2)
```

See also: section \@ref(block-param) and \@ref(block-theta).

Notes:

* Multiple blocks are allowed
* Values are evaluated by the `R` interpreter


### `$CMT`  and  `$INIT`
Declare the names of all compartments in the model. 

* For `$CMT` give the names of compartments; initial values are assumed to be 0
* For `$INIT` give the name and initial value for all compartments

Examples:
```{c,eval=FALSE}
$CMT GUT CENT RESPONSE
```

```{c,eval=FALSE}
$INIT GUT  = 0, CENT = 0, RESPONSE = 25
```

Annotated examples:
```{c, eval=FALSE}
$CMT @annotated
GUT      : Dosing compartment (mg)
CENT     : Central PK compartment (mg)
RESPONSE : Response
```

```{c, eval=FALSE}
$INIT @annotated
GUT      :   0 : Dosing compartment (mg)
CENT     :   0 : Central PK compartment (mg)
RESPONSE :  25 : Response
```



See `?init` in the `R` help system after loading `mrgsolve`.

### `$MAIN`  \ {#block-main}
This code block has two main purposes:

* Derive new algebraic relationships between parameters, random, effects and other derived variables
* Set the initial conditions for model compartments

For users who are familiar with NONMEM, `$MAIN` is similar to `$PK`.

`$MAIN` is wrapped into a `C++` function and compiled / loaded by `mrgsolve`.  
The `MAIN` function gets called just prior to advancing the system from the 
current time to the next time for each record in the data set.  `$MAIN` also 
gets called several times before starting the problem (`NEWIND == 0`) and 
just prior to simulating each individual (`NEWIND == 1`).  Finally, `$MAIN` 
gets called every time the model initial conditions are queried with `init()`.  

New variables may be declared in `$MAIN`.  See section \@ref(new-variables) 
for details.

Examples:

```{c,eval=FALSE}
$CMT CENT RESP

$PARAM KIN = 100, KOUT = 2, CL = 1, VC = 20

$MAIN

RESP_0 = KIN/KOUT;

double ke = CL/VC;

```


### `$ODE` \ {#block-ode}

Use `$ODE` to define model differential equations.  For all compartments assign 
the value of the differential equation to `dxdt_CMT` where `CMT` is the name 
of the compartment.  The `dxdt_` equation may be a function of model 
parameters (via `$PARAM`), the current value of any compartment (`CMT`) 
or any user-derived variable. 

For example:
```{c,eval=FALSE}
$CMT GUT CENT

$ODE
dxdt_GUT = -KA*GUT;
dxdt_CENT = KA*GUT - KE*CENT;

```

It is important to make sure that there is a `dxdt_` expression defined for 
every compartment listed in `$CMT` or `$INIT`, even if it is `dxdt_CMT = 0;`

The `$ODE` function is called repeatedly during a simulation run.  So it is 
wise to do as many calculations as possible outside of `$ODE`, usually in 
`$MAIN`.  But remember that any calculation that depends on an amount in a
compartment and helps determine the `dxdt_` expression in a model must be 
written in `$ODE`.


New variables may be declared in `$ODE`.  See section \@ref(new-variables) 
for details.

For example:
```{c,eval=FALSE}
$CMT CENT RESP
$PARAM VC = 100, KE = 0.2, KOUT = 2, KIN = 100
$ODE
double CP = CENT/VC;
double INH = CP/(IMAX+CP)

dxdt_CENT = -KE*CENT;
dxdt_RESP =  KIN*(1 - INH) - RESP*KOUT;

```

If the model needs to refer to the current time, use the `SOLVERTIME` variable.

Notes:

* `$ODE` is written in `C++` syntax; every line must end in `;`
* There may be only one `$ODE` block in a model


### `$TABLE` \ {#block-table}
Use `$TABLE` to interact with parameters, compartment values, and other user-defined variables __after__ the system advances to the next time.  

For example:
```{c,eval=FALSE}
$TABLE
double CP = CENT/VC;
```

__NOTE__
`mrgsolve` formerly had a `table()` macro for inserting derived values into simulated output. This macro
has been deprecated.  The only way to insert derived values into the simulated output is via `$CAPTURE`.

__NOTE__
When variables are marked for capture 
(see \@ref(block-capture)), the values 
of those variables are saved at the __end__
of the `$TABLE` function.  This process is carried
out automatically by `mrgsolve` and therefore
requires no user intervention. 


### `$PREAMBLE` \ {#block-preamble}
This is the fourth C++ code block.  It is called once in two different settings:

1. Immediately prior to starting the simulation run
1. Immediately prior to calling `$MAIN` when calculating initial conditions


`$PREAMBLE` is a function that allows you to set up your C++ environment.  It is only called 
one time during the simulation run (right at the start).  The code in this block is typically 
used to configure or initialize C++ variables or data structures that were declared in `$GLOBAL`. 

For example:

```{c,eval=FALSE}
$PLUGIN Rcpp

$GLOBAL 
namespace{
  Rcpp::NumericVector x;
}

$PREAMBLE
x.push_back(1);
x.push_back(2);
x.push_back(3);

$MAIN 
<some code that uses x vector>
```

In this example, we want to use a numeric vector `x` and declare it in `$GLOBAL` so that we
can use it anywhere else in the code (the declaration is also made in an unnamed namespace to 
ensure that the variable is local to the model file).  Then, in `$PREAMBLE`, we put 3 numbers
into the vector and we use `x` in `$MAIN`.  Since `$MAIN`, `$TABLE` and (especially) `$ODE` are called repeatedly as the simulation run procees, we put the initialization of `x` in `$PREAMBLE`
to make sure the initialization of `x` only happens once.

Notes:

* `$PREAMBLE` is written in `C++` syntax; every line must end in `;`
* There may be only one `$PREAMBLE` block in a model
* Like `$MAIN`, `$ODE` and `$TABLE`, `double`, `int` and `bool` variables initialized in `$PREAMBLE` are actually initialized for global (within the model file)

See also: \@ref(block-plugin).

### `$CAPTURE` \ {#block-capture}
This is a block to identify variables that should be captured in the simulated output. 

For example:
```{c,eval=FALSE}
$PARAM A = 1, B = 2

$MAIN
double C = 3;
bool yes = true;

$CAPTURE A B C yes
```

This construct will result in 4 additional columns in the simulated output 
with names `A`, `B`, `C`, and `yes`.  

Users can also rename captured variables by providing a 
`newname = oldname` specification. 

```{c, eval = FALSE}
$PARAM WT = 70, THETA1 = 2.2

$MAIN
double CL = THETA1*pow(WT/70,0.75)*exp(ETA(1));

$OMEGA 1

$CAPTURE WEIGHT = WT TVCL = THETA2 CL  ETA(1)
```

In this example, the names of the captured data items
will be `WEIGHT,TVCL,CL,ETA_1`. 

Users can use the `capture` type to declare variables in `$MAIN` and `$TABLE`. 
`capture` types are really `double`s, but using that type will signal `mrgsolve`
to automatically capture that value.  For example:

```{c,eval=FALSE}
$PARAM VC = 300

$CMT CENT

$TABLE
capture DV = (CENT/VC);
```

Since we used type `capture` for `DV`, `DV` will show up as a column in the
simulated data.

Annotated example:

```{c,eval=FALSE}
$MAIN
double CLi = TVCL*exp(ECL);

$TABLE
double DV = (CENT/VC)*exp(PROP);

$CAPTURE @annotated
CLi : Individual clearance (L/hr)
DV  : Plasma concentration (mcg/ml)
```




New variables may be declared in `$TABLE`.  See section \@ref(new-variables) for details.


### `$OMEGA` \ {#block-omega}
See `?modMATRIX` for more details about options for this block.

Use this block to enter variance/covariance matrices for subject-level random effects drawn from multivariate normal distribution. All 
random effects are assumed to have mean of 0.  Off diagonal elements for block matrices are assumed to be correlation coefficients if the 
`@correlation` option is used (see below).

By default, a __diagonal__ matrix is assumed.  So:

```{c, eval=FALSE}
$OMEGA
1 2 3
```
will generate a 3x3 omega matrix.

A __block__ matrix may be entered by using `block=TRUE`. So:
```{c, eval=FALSE}
$OMEGA @block
0.1 0.02 0.3
```
will generate a 2x2 matrix with covariance 0.02.

A 2x2 matrix where the off-diagonal element is a correlation, not a covariance can be specified like this:
```{c,eval=FALSE}
$OMEGA @correlation
0.1 0.67 0.3
```

Here, the correlation is 0.67. `mrgsolve` will calculate the covariances and substitute these values.  The matrix will be stored and used with these covariances, not the correlation.

A name can be assigned to each matrix:
```{c, eval=FALSE}
$OMEGA @name PK @block
0.2 0.02 0.3

$OMEGA @name PD
0.1 0.2 0.3 0.5
```
to distinguish between multiple `$OMEGA` blocks and to facilitate updating later. The model in the preceeding example will have two 
`$OMEGA` matrices: 2x2 and 4x4.


Annotated example (diagonal matrix):
```{c, eval=FALSE}
$OMEGA @annotated
ECL: 0.09 : ETA on clearance
EVC: 0.19 : ETA on volume
EKA: 0.45 : ETA on absorption rate constant
```

Annotated example (block matrix):
```{c, eval=FALSE}
$OMEGA @annotated @block
ECL: 0.09 : ETA on clearance
EVC: 0.001 0.19 : ETA on volume
EKA: 0.001 0.001 0.45 : ETA on absorption rate constant
```


Notes:

* Multiple `$OMEGA` blocks are allowed


### `$SIGMA` \ {#block-sigma}
See `?modMATRIX` for more details about options for this block.

Use this block to enter variance/covariance matrices for within-subject random effects drawn from multivariate normal distribution. All 
random effects are assumed to have mean of 0.  Off diagonal elements for block matrices are assumed to be correlation coefficients if the 
`@correlation` option is used (see below).

The `$SIGMA` block functions like the `$OMEGA` block.  See `$OMEGA` for details. 

### `$SET`
Use this code block to set different options for the simulation. Use a `name=value` format, where `value` is evaluated by the `R` interpreter.  

Most of the options that can be entered in `$SET` are passed to `update`. 

For example:
```{c,eval=FALSE}
$SET end = 240, delta=0.5, req=s(RESP)
```

Here, we set the simulation `end` time to 240, set the time difference between two
adjacent time points to 0.25 time units, and request only the `RESP`onse compartment in the simulated output.


### `$GLOBAL`
The `$GLOBAL` block is for writing `C++` code that is outside of `$MAIN`, `$ODE`, and `$TABLE`.

There are no artificial limit on what sort of `C++` code can go in `$GLOBAL`.  However there are two more-common uses:

1. Write `#define` preprocessor statements
1. Define global variables, usually variables other than `double`, `bool`, `int` (see \@ref(new-variables))


__Preprocessor directives__
Preprocessor  `#define` directives are direct substibutions that the `C++` preprocessor makes prior to compiling your code.  

For example:
```{c,eval=FALSE}
$GLOBAL
#define CP (CENT/VC)
```
When this preprocessor directive is included, everywhere the preprocessor finds a `CP` token it will substitute `(CENT/VC)`.  Both `CENT` and `VC` must be defined and the ratio of `CENT` to `VC` will be calculated depending on whatever the current values are.  Notice that we included parentheses around `(CENT/VC)`.  This makes sure the ratio between the two is taken first, before any other operations involving `CP`.  

__Declaring global variables__
Sometimes, you may wish to use global variables and have more control over how they get declared.  

```{c,eval=FALSE}
$GLOBAL
bool cure = false;
```
With this construct, the boolean variable `cure` is declared and defined right as the model is compiled.  


### `$PKMODEL`
This code block implements a one- or two-compartment PK model where the system is calculated by algebraic equations, not ODEs.  `mrgsolve` handles the calculations and an error is generated if both `$PKMODEL` and `$ODE` blocks are included in the same model specification file.

This is an options-only block. The user must specify the number of compartments (1 or 2) to use in the model as well as whether or not to include a depot dosing compartment.  See `?PKMODEL` for more details about this block, including specific requirements for symbols that must be defined in the model specification file.

The `$CMT` or `$INIT` block must also be included with an appropriate number of compartments.  Compartment names, however, may be determined by the user.

Example:
```{c, eval=FALSE}
$CMT GUT CENT PERIPH
$PKMODEL ncmt=2, depot=TRUE
```

As of version `0.8.2`, we can alternatively specify the compartments
right in the `$PKMODEL` block:
```{c,eval=FALSE}
$PKMODEL cmt="GUT CENT PERIPH", depot = TRUE
```
Specifying three compartments with `depot=TRUE` implies
`ncmt=2`.  Notice that a separate `$CMT` block is not 
appropriate when `cmt` is specified in `$PKMODEL`.


### `$PLUGIN` \ {#block-plugin}
Plugins are a way to add extensions to your `mrgsolve` model.  When a plugin is 
recruited into the model, `mrgsolve` may do one or more of the following:

- Link to another `R` package during compilation, including
    - `Rcpp` to allow you to write `Rcpp` code in your specification
    - `mrgx` to provide some extra C++ functions (see below)
    - `RcppArmadillo` to both `Rcpp` and `Armadillo` in your specification
    - `BH` to use `boost` headers
- Include appropriate header files during compilation
    - For example, when the `Rcpp` plugin is called, `mrgsolve` will `#include <Rcpp.h> at the top of your model file

You recruit a plugin in the `$PLUGIN` block

```{c, eval=FALSE}
$PLUGIN Rcpp
```

The example above will bring `Rcpp` headers into the model code.  

Plugins that you can use:
- `Rcpp` to write any `Rcpp` code
- `mrgx` additional functions provided by `mrgsolve` that help use `Rcpp` in your model
- `RcppArmadillo` to write any `Rcpp` or `RcppArmadillo` code
- `BH` to link to `boost` headers

Note that `Rcpp`, `RcppArmadillo` and `BH` only allow you to link to those headers.  To take advantage of that, you will need to know how to use `Rcpp`, `boost` etc.  For the `BH` plugin, no headers are included for you; you must include the proper headers you want to use in `$GLOBAL`.


__Rcpp__ example

```{c,eval=FALSE}

$PLUGIN Rcpp

$MAIN

if(NEWIND <=1) {
  double wt = R::rnorm(70,20);
  double sex = R::rbinom(1,0.51);
}

```

__mrgx__

For example

```{c,eval=FALSE}

$PLUGIN Rcpp mrgx

```

Functions provided by `mrgx`:

  -  `T get<T>(std::string <pkgname>, std::string <objectname>)`
      - This gets an object of any Rcpp-representable type (`T`) from any package
  -  `T get<T>(std::string <objectname)`
      - This gets an object of any Rcpp-representable type (`T`) from `.GlobalEnv`
  -  `T get<T>(std::string <objectname>, databox& self)`
      - This gets an object of any Rcpp-representable type (`T`) from `$ENV`
  -  `double rnorm(double mean, double sd, double min, double max)`
      - Simulate one variate from a normal distribution that is between `min` and `max`
  -  `double rlognorm(double mean, double sd, double min, double max)`
      - Same as `mrgx::rnorm`, but the simulated value is passed to `exp` after simulating
  -  `Rcpp::Function mt_fun()` 
      - Returns `mrgsolve::mt_fun`; this is usually used when declaring a `R` function in `$GLOBAL`
      - Example: `Rcpp::Function print = mrgx::mt_fun();`


__IMPORTANT__ All of these functions are in the `mrgx` namespace.  So, in order to call these 
functions you must include `mrgx::` namespace identifier to the front of the function name.  For example, don't use `rnorm(50,20,40,140)`; use `mrgx::rnorm(50,20,40,140)`.


__Get a numeric vector from `$ENV`__
```{c,eval=FALSE}
$PLUGIN Rcpp mrgx

$ENV
x <- c(1,2,3,4,5)

$GLOBAL
Rcpp::NumericVector x;

$PREAMBLE
x = mrgx::get<Rcpp::NumericVector>("x", self);

```


__Get the `print` function from `package:base`__
```{c,eval=FALSE}
$PLUGIN Rcpp mrgx

$GLOBAL
Rcpp::Function print = mrgx::mt_fun();

$PREAMBLE
print = mrgx::get<Rcpp::Function>("base", "print");

$MAIN
print(self.rown);
```

Note that we declare the `print` in `$GLOBAL` and use the `mt_fun()` place holder.

__Simulate truncated normal variables__
This simulates a weight that has mean 80, standard deviation 20 and is greater than 
40 and less than 140.

```{c,eval=FALSE}
$PLUGIN Rcpp mrgx

$MAIN
if(NEWIND <=1) {
  double WT = mrgx::rnorm(80,20,40,140);
}
```

See also: \@ref(block-preamble).

### `$THETA`  \ {#block-theta}
Use this code block as an efficient way to add to the parameter list where 
names are determined by a prefix and a number.  By default, the prefix is `THETA` and the number sequentially numbers the input values. 

For example:

```{c,eval=FALSE}
$THETA
0.1 0.2 0.3
```
is equivalent to
```{c,eval=FALSE}
$PARAM THETA1 = 0.1, THETA2 = 0.2, THETA3 = 0.3
```

Annotated example:

```{c,eval=FALSE}
$THETA @annotated
0.1 : Typical value of clearance (L/hr)
0.2 : Typical value of volume (L)
0.3 : Typical value of ka (1/hr)
```

To change the prefix, use `@name` directive

```{c,eval = FALSE}
$THETA @name theta
0.1 0.2 0.3
```

would be equivalent to
```{c,eval = FALSE}
$PARAM theta1 = 0.1, theta2 = 0.2, theta3 = 0.3
```


See also: \@ref(block-param) .


### `$NMXML`  \ {#block-nmxml}
The `$NMXML` block lets you read and incorporate results from a NONMEM run into 
your `mrgsolve` model.  From the NONMEM run, `THETA` will be imported into your
parameter list (see \@ref(block-param) and \@ref(component-param)), `OMEGA` will 
be captured as an `$OMEGA` block (\@ref(block-omega)) and `SIGMA` will be 
captured as a `$SIGMA` block (\@ref(block-sigma)).  Users may optionally omit 
any one of these from being imported.

`$NMXML` contains a `project` argument and a `run` argument.  By default, the 
estimates are read from from the file `project/run/run.xml`.  That is, it is 
assumed that there is a directory named `run` that is inside the `project` 
directory where `$NMXML` will find `run.xml`.  Your NONMEM run directories may
not be organized in a way that is compatible with this default.  In that case, 
you will need to provide the `file` argument, which should be the path to the 
`run.xml` file, either as a full path or as a path relative to the current 
working directory.

For help on the arguments / options for `$NMXML`, please see the `?nmxml` 
help topic in your `R` session after loading the `mrgsolve` package.

__An example__

There is a NONMEM run embedded in the `mrgsolve` package 
```{r}
path <- file.path(path.package("mrgsolve"),"nonmem")
list.files(path, recursive=TRUE)

```

 
We can create a `mrgsolve` control stream that will import `THETA`, `OMEGA` and 
`SIGMA` from that run using the `$NMXML` code block.

```{r}
code <- '
$NMXML
run = 1005
project = path
olabels = s_(ECL, EVC, EKA)
slabels = s_(PROP, ADD)

$MAIN
double CL = THETA1*exp(ECL);
double V2 = THETA2*exp(EVC);
double KA = THETA3*exp(EKA);
double Q = THETA4;
double V3 = THETA5;

$PKMODEL ncmt=2, depot=TRUE

$CMT GUT CENT PERIPH

$TABLE
double CP = (CENT/V2)*(1+PROP) + ADD/5;

$CAPTURE CP

$SET delta=4, end=96
'
```


__NOTE__: in order to use this code, we need to install the `XML` package.

```{r}
mod <- mcode("nmxml", code, quiet=TRUE)

mod

param(mod)

revar(mod)

```


An infusion into `GUT`:
```{r}
set.seed(2922)
#+
mod %>%
  ev(amt=1000, rate=1000/36, cmt="GUT") %>%
  mrgsim(nid=10) %>%
  plot
```

### `$INCLUDE`  \ {#block-include}
To include your own header file(s) in a model use `$INCLUDE`
```{c,eval=FALSE}
$INCLUDE
mystuff.h
otherstuff.h
```

or

```{c,eval=FALSE}
$INCLUDE
mystuff.h otherstuff.h
```

or

```{c,eval=FALSE}
$INCLUDE
mystuff.h,  otherstuff.h
```


`mrgsolve` will insert proper `#include` preprocessor directives into the `C++` code that gets compiled.

__Requirements__

  - All header files listed in `$INCLUDE` are assumed (expected) to be under the `project` directory; don't use `$INCLUDE` for header files that are in any other location
  - An error is generated if the header file does not exist
  - An error is generated if any quotation marks are found in the file name (don't use quotes around the file name; `mrgsolve` will take care of that)
  - A warning is issued if the header file does not end in `.h`
  - When the header file is changed (MD5 checksum changes), the model will be forced to be rebuilt (recompiled) when `mread` or `mcode` is called; this feature is only available for header files listed in `$INCLUDE` (see below)
  - Do not use `$INCLUDE` to include `Rcpp`, `boost`, `RcppArmadillo` or `RcppEigen` headers; use the appropriate `$PLUGIN` instead

For applications that don't fit into the requirements listed above, users can always include header files in the model in `$GLOBAL` like this:

```{c,eval=FALSE}
$GLOBAL
#include "/Users/me/libs/mystuff.h"
```

But be careful when doing this: if there are changes to `mystuff.h` but not to any other part of the model specification, the model may not be fully compiled when calling `mread`.  In this case, always use `preclean=TRUE` argument to `mread` to force the model to be built when calling `mread`.


### `$ENV` \ {#block-env}
This block is all `R` code (just as you would code in a stand-alone `R` script. The code is parsed and evaluated into a new environment when the model is compiled.  Objects inside `$ENV` can be utilized in different C++ functions (see \@ref(block-plugin)) or other parts of the simulation process.

For example:

```{c,eval=FALSE}
$ENV

Sigma <- cmat(1,0.6,2)

mu <- c(2,4)

cama <- function(mod) {
  mod %>%
    ev(amt=100, ii=12, addl=10) %>% 
    mrgsim(obsonly=TRUE,end=120)
}

```

## Variables and Macros

This section describes some macros and internal variables that can 
be used in model specification files.  It should be clear from the usage 
examples which variables can be set by the user and which are to be 
read or checked.  All internal variables are pre-defined and pre-initialized.
The user should never try to declare an internal variable; this will 
always result in an compile-time error. 

In the following section, we adopt the convention that `CMT` stands for a compartment in the model.

### `ID`
The current subject identifier. `ID` is an alias for `self.id`.

### `TIME`
Gives the time in the curent data set record.  This is usually only used in `$MAIN` or `$TABLE`.  `TIME` is an alias for `self.time`.  Contrast with `SOLVERTIME`.

### `SOLVERTIME`
Gives the time of the current timestep taken by the solver.  This is can only be used in `$ODE`.  Contrast with `TIME`.

### `EVID`
`EVID` is an event id indicator.  `mrgsolve` recognized the following 
event IDs:

* 0 = an observation record
* 1 = a bolus or infusion dose
* 2 = other type event
* 3 = system reset
* 4 = system reset and dose
* 8 = replace

`EVID` is an alias for `self.evid`.

### `NEWIND`
`NEWIND` is a new individual indicator, taking the following values:

* 0 for the first event record of the data set
* 1 for the first event record of a subsequent individual
* 2 for subsequent event record for an individual

For example:
```{c, eval=FALSE}
$GLOBAL 
int counter = 0;

$MAIN
if(NEWIND <=1) {
  counter  = 0;
}
```

`NEWIND` is an alias for `self.newind`.

### `self.cmt`
The current compartment number regardless of whether it was given as `cmt` or 
`CMT` in the data set.  There is no alias for `self.cmt`.

For example:

```{c,eval=FALSE}

$TABLE
double DV = CENT/VC + EPS(1);
if(self.cmt==3) DV = RESPOSE + EPS(2);

```


### `self.amt`
The current `amt` value regardless of whether it was given as `amt` or `AMT` 
in the data set.  There is no alias for `self.amt`.

### `self.nid`
The number of IDs in the data set.

### `self.idn`
The current id number.  Numbers start at 0 and increase by one to `self.nid-1`. 

### `self.nrow`
The number of rows in the output data set.

### `self.rown`
The current row number.  Numbers start at 0 and increase by one to `self.rown-1`. 


### `ETA(n)`
`ETA(n)` is the value of the subject-level variate drawn from the model 
`OMEGA` matrix.  `ETA(1)` through `ETA(25)` have default values of zero so 
they may be used in a model even if appropriate `OMEGA` matrices have not been
provided. 

For example:
```{c, eval=FALSE}
$OMEGA
1 2 3

$MAIN
double CL = TVCL*exp(ETA(1));
double VC = TVVC*exp(ETA(2));
double KA = TVKA*exp(ETA(3));
```
Here, we have a 3x3 `OMEGA` matrix. `ETA(1)`, `ETA(2)`, and `ETA(3)` will be
populated with variates drawn from this matrix.  `ETA(4)` through `ETA(25)` 
will be populated with zero. 

### `EPS(n)`
`EPS(n)` holds the current value of the observation-level random variates drawn
from `SIGMA`.  The basic setup is the same as detailed in `ETA(n)`.

Example:
```{c,eval=FALSE}
$CMT CENT

$PARAM CL=1, VC=20

$SIGMA
labels=s(ADD,PROP)
25 0.0025

$TABLE
double DV = (CENT/VC)*(1+PROP) + ADD;

```



### `table(name) `
This macro has been deprecated.  Users should __not__ use code like this:

```{c,eval=FALSE}
$TABLE
table(CP) = CENT/VC;
```

But rather this:
```{c,eval=FALSE}
$TABLE 
double CP = CENT/VC;

$CAPTURE CP
```

See: section \@ref(block-table) and also \@ref(block-capture)


### `F_CMT` \ {#spec-f}
For the `CMT` compartment, sets the bioavability fraction for that compartment. 

Example:
```{c,eval=FALSE}
$MAIN
F_CENT = 0.7;
```


### `ALAG_CMT` \ {#spec-alag}
For the `CMT` compartment, sets the lag time for doses into that compartment. 

Example:
```{c,eval=FALSE}
$MAIN
ALAG_CENT = 0.25;
```


### `R_CMT` \ {#spec-r}
For the `CMT` compartment, sets the infusion rate for that compartment.  The
infusion rate is only set via `R_CMT` when `rate` in the data set or event 
object is set to `-1`. 

Example:
```{c,eval=FALSE}
$MAIN
R_CENT = 100;
```


### `D_CMT` \ {#spec-d}
For the `CMT` compartment, sets the infusion duration for that compartment.  
The infusion duration is only set via `D_CMT` when `rate` in the data set or 
event object is set to `-2`. 

Example:
```{c,eval=FALSE}
$MAIN
D_CENT = 2;
```


## Derive new variables {#new-variables}
New `C++` variables may be derived in `$GLOBAL`, `$MAIN`, `$ODE` and `$TABLE`.
Because these are `C++` variables, the type of variable being used must be
declared.  For the vast majority of applications, the `double` type is used
(double-precision numeric value). 

```{c,eval=FALSE}
$MAIN
double CLi = TVCL*exp(ETA(1));
```

We want `CLi` to be a numeric value, so we use `double`.  To derived a `boolean` variable, write
```{c,eval=FALSE}
$MAIN
bool cure = false;
```

When variables of the type `double`, `int`, and `bool` are declared and 
initialized in `$MAIN`, `$ODE`, `$TABLE`, `mrgsolve` will detect those
declarations, and modify the code so that the variables are actually declared 
in `$GLOBAL` not in `$MAIN`, `$ODE`, or `$TABLE`.  This is done so that 
variables declared in one code block can be read and modified in another 
code block.  

For example, the following code:
```{c,eval=FALSE}
$MAIN
double CLi = TVCL*exp(ETA(1));
```

gets translated to:
```{c,eval=FALSE}
$GLOBAL
double CLi;

$MAIN
CLi = TVCL*exp(ETA(1));
```

This way, we can still read `CLi` in `$TABLE`:
```{c,eval=FALSE}
$MAIN
double CLi = TVCL*exp(ETA(1));
double VCi = TVVC*exp(ETA(2));

$TABLE
double KEi = CLi/VCi;

$CAPTURE KEi
```

To declare a variable that is local to a particular code block:

```{c,eval=FALSE}
$MAIN
localdouble CLi = TVCL*exp(ETA(1));

```


## Examples
The following sections show example model specification.  The intention is to 
show how the different blocks, macros and variables can work together to make 
a functional model.  Some models are given purely for illustrative purpose and 
may not be particularly useful in application.

### Simple PK model

Notes:

* Basic PK parameters are declared in `$PARAM`; every parameter needs to be
assigned a value
* Two compartments `GUT` and `CENT` are declared in `$CMT`; using `$CMT` 
assumes that both compartments start with 0 mass
* Because we declared `GUT` and `CENT` as compartments, we write `dxdt_` 
equations for both in `$ODE`
* In `$ODE`, we refer to parameters (`CL/VC/KA`) and the amounts in each
compartment at any particular time (`GUT` and `CENT`)
* `$ODE` should be `C++` code; each line ends in `;`
* We derive a variable called `CP` in `$TABLE` and `$CAPTURE` that value 
so that it appears in the simulated output

```{c, eval=FALSE}
$PARAM CL = 1, VC = 30, KA = 1.3

$CMT GUT CENT

$ODE

dxdt_GUT = -KA*GUT;
dxdt_CENT = KA*GUT - (CL/VC)*CENT;

$TABLE
double CP = CENT/VC;

$CAPTURE CP
```


### PK/PD model

Notes:

* We use a preprocessor `#define` directive in `$GLOBAL`; everywhere in the 
model where a `CP` token is found, the expression `(CENT/VC)` ... with 
parentheses ... is inserted
* We write the initial value for the `RESP` compartment in `$MAIN` as a 
function of two parameters `KIN/KOUT`
* A new variable - `INH`- is declared and used in `$ODE`
* Since `CP` is defined as `CENT/VC`, we can "capture" that name/value in
`$CAPTURE`
* Both `$MAIN` and `$ODE` are `C++` code blocks; don't forget to add the `;` 
at the end of each statement

```{c, eval=FALSE}
$PARAM CL = 1, VC = 30, KA = 1.3
KIN = 100, KOUT = 2, IC50 = 2

$GLOBAL
#define CP (CENT/VC)


$CMT GUT CENT RESP

$MAIN
RESP_0 = KIN/KOUT;

$ODE

double INH = CP/(IC50+CP);

dxdt_GUT = -KA*GUT;
dxdt_CENT = KA*GUT - (CL/VC)*CENT;
dxdt_RESP = KIN*(1-INH) - KOUT*CENT;


$CAPTURE CP

```

### Population PK model with covariates and IOV

Notes:

* Use `$SET` to set the simulation time grid from 0 to 240 by 0.1
* There are two `$OMEGA` matrices; we name them `IIV` and `IOV`
* The IIV "etas" are labeled as `ECL/EVC/EKA`; these are aliases to
`ETA(1)/ETA(2)/ETA(3)`.  The `IOV` matrix is unlabeled; we must refer to
`ETA(4)/ETA(5)` for this
* Because `ETA(1)` and `ETA(2)` are labeled, we can "capture" them as `ECL` and
`EVC`
* We added zeros for both `$OMEGA` matrices; all the etas will be zero until we
populate those matrices (section \@ref(topic-matrix-update))


```{c,eval=FALSE}
$PARAM TVCL = 1.3, TVVC=28, TVKA=0.6, WT=70, OCC=1

$SET delta=0.1, end=240

$CMT GUT CENT

$MAIN

double IOV = IOV1
if(OCC==2) IOV = IOV2;

double CLi = exp(log(TVCL) + 0.75*log(WT/70) + ECL + IOV);
double VCi = exp(log(TVVC) + EVC;
double KAi = exp(log(TVKA) + EKA;

$OMEGA @name IIV @labels ECL EVC EKA
0 0 0
$OMEGA @name IOV @labels IOV1 IOV2
0 0

$SIGMA 0

$ODE
dxdt_GUT = -KAi*GUT;
dxdt_CENT = KAi*GUT - (CLi/VCi)*CENT;

$TABLE
double CP = CENT/VCi;

$CAPTURE IOV ECL EVC CP
```




