---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Steady state {#steady-state}

```{r, setup,  echo=FALSE,message=FALSE}
source("setup.R")
```

## Introduction

Within mrgsolve, the term "steady state" (`SS`) applies specifically to the 
pharmacokinetic dosing system and indicates that the rate of drug administration
is equal to the rate of drug elimination. Steady state dosing can take the form
of repeated intermittent doses (bolus or infusion, administered intermittently
at a given dosing interval) or a continuous infusion administered to steady
state.

### Intermittent doses

The user can direct mrgsolve to advance the system to steady state for
intermittent dosing by including `ss=1` in an event object or input data set.
For example:

```{r}
mod <- modlib("pk1", end = 120)

dose <- ev(amt = 100, ii = 12, addl = 4, ss = 1)

mrgsim(mod, dose, recsort = 3) %>% plot()
```

In this example, the `ss=1` flag tells mrgsolve to advance the system to steady
state under a dosing regimen of 100 mg every 12 hours (and then give a total of
5 doses). When using the `ss=1` flag, the user is required to indicate the 
dosing interval (here every 12 hours) and additional doses are optional. Similar
behavior can be achieved for intermittent infusions by setting the infusion 
`rate`.

### Continuous infusion

A continuous infusion can be dosed to steady state by including the `ss=1` flag, 
a value for `rate` (any positive rate or -1 if the rate is being modeled), and 
setting the dose amount (`amt`/`AMT`) to zero:

```{r}
infus <- ev(amt = 0, rate = 5, ss = 1, cmt = "CENT")

mrgsim(mod, infus, recsort = 3) %>% plot()
```

Because `CL` is equal to 1 in this model, we see that the continuous 
(never-ending) infusion was started at steady state with a value of 5.


## Advance to SS

It's important to recognize that `SS` is related to the PK dosing system; it is
finding the state of the system after an infinite number of doses have been
administered under a certain regimen.  And this is essentially how mrgsolve goes
about finding steady state: when the `ss=1` flag is encountered, mrgsolve starts
repeatedly administering doses and advancing the system to the next dose
according to the inter-dose interval (`ii`). Once mrgsolve determines that the
amounts in the system at any dose are the same as they were at the preceding
dose, mrgsolve declares that `SS` has been achieved.

## Control advance to SS

### Tolerances for SS

mrgsolve uses a local error estimate to determine the degree to which
concentrations are changing or not changing between doses on the way to `SS`.
This is determined by a relative tolerance parameter (`ss_rtol`) and an absolute
tolerance parameter (`ss_atol`). As of mrgsolve version 0.10.3, these tolerances
are distinct from the tolerances used for solving the differential equations
(`rtol` and `atol`, respectively). Note that when advancing to `SS` in an ODE
model, `ss_rtol` (the relative tolerance for determining `SS`) must be larger
(less precise) than `rtol` (the relative tolerance used by the ODE solver). Once
the difference between two trough concentrations is less than A$_{trough}$ *
`ss_rtol` + `ss_atol`, then the system is said to be at steady state.  By
default, this calculation is done for every single compartment in the model and
all compartments have to meet this criteria before the system is said to be at
steady state.  So, increasing `ss_rtol` (say from 1e-8 to 1e-3) will also allow
us to call it "good" with respect to steady state sooner. 

Both tolerances for steady state are stored in the model object and can be set
with the update method.  For example, 

```{r, eval=FALSE}
mod <- house()

mod <- update(mod, ss_rtol = 1e-5, ss_atol = 1e-8)
```

### Max dose number

It was noted above that mrgsolve advances the system to steady state with a
brute force approach: doses are repeatedly administered at a regular interval
(for intermittent `SS`) until pre-dose concentrations are the same dose to dose.
mrgsolve sets an upper limit (equal to `r formals(do_mrgsim)$ss_n`) to the
number of doses that will be administered before giving up on trying to find
steady state. Once this maximum number of doses is exceeded, mrgsolve will 
issue a warning that steady state was not achieved and continue on with the 
problem. For example:

```{r}
dose <- ev(amt = 100, ii = 12, ss = 1, VC = 800)

out <- mrgsim(house(), dose)
```

Here, mrgsolve administered 500 doses and the pre-dose concentrations were still
not similar enough to declare the system to be at steady state.

The maximum dose number can be set with the `ss_n` argument to `mrgsim` (or one 
of the variants; see `?do_mrgsim` help topic). This number can be increased to 
prevent the warning:

```{r}
out <- mrgsim(house(), dose, ss_n  = 1000)
```

Alternatively, the `ss_fixed` argument to mrgsim (see `?do_mrgsim`) can be 
set to `TRUE` to silence the warning. In this case, **up to** `ss_n` doses will 
be administered and if `SS` is not achieved with that many doses, the problem
will continue with no warning.

### Include / exclude compartments for SS

Sometimes it might be sufficient to only consider one compartment when 
determining `SS` (e.g. the central compartment in a PK model). Other times, 
it might be helpful to exclude a compartment when determining `SS` (e.g. a 
depot dosing compartment when concentrations can get very small toward the 
end of the dosing interval).

mrgsolve allows the user to identify compartments to include or exclude in 
determining `SS`.  This is done through the `ss_cmt` option in `$SET`.  To 
only consider the `CENT` compartment, write the following in the model file:

```{c, eval=FALSE}
[ set ] ss_cmt = "CENT"
```

This says to **only** look at the `CENT` compartment when determining `SS`.

Alternatively, you can exclude certain compartments like this:

```{c, eval=FALSE}
[ set ] ss_cmt = "-GUT,DEPOT"
```

This says to ignore the `GUT` and `DEPOT` compartments when determining `SS`.

As another example, you might want to exclude an accumulator compartment when 
calculating `SS`

```{c, eval=FALSE}
[ set ] ss_cmt = "-AUC"

[ ode ] 
dxdt_CENT = -kel * CENT;

dxdt_AUC = CENT/VC;
```

This is just a partial model snippet, but it shows how you might exclude the 
AUC compartment when determining `SS`. 

### SS_ADVANCE flag

mrgsolve also provides an `SS_ADVANCE` indicator that is passed into `$ODE` and 
evaluates to `true` when the system is being advanced to steady state. So a
better way to exclude the accumulator compartment from being considered for 
`SS` calculation would be this:

```{c eval=FALSE}
[ ode ] 

dxdt_AUC = CENT/VC;

if(SS_ADVANCE) dxdt_AUC = 0;
```

This code prevents the `AUC` compartment from changing during the advance to 
`SS` and the dose to dose difference in `AUC` will always be zero, effectively 
excluding this compartment from factoring into the `SS` determination. This 
should be the preferred approach to dealing with an `AUC` compartment. 
