```{r,echo=FALSE,message=FALSE}
source("setup.R")
```


# Topics {#topics}

## Annotated model specification {#topic-annotated}
Here is a complete annotated `mrgsolve` model.  The 
goal was to get in several of the most common
blocks that you might want to annotate.  The different 
code blocks are rendered here separately for clarity in
presentation; but users should include all relevant
blocks in a single file (or R string).


```{r,eval=FALSE}
$PROB

# Final PK model

- Author: Pmetrics Scientist
- Client: Pharmaco, Inc.
- Date: `r Sys.Date()`
- NONMEM Run: 12345
- Structure: one compartment, first order absorption
- Implementation: closed form solutions
- Error model: Additive + proportional
- Covariates:
  - WT on clearance
- SEX on volume
- Random effects on: `CL`, `V`, `KA`
```

```{r,eval=FALSE}
[PARAM] @annotated
TVCL : 1.1   : Clearance (L/hr)
TVV  : 35.6  : Volume of distribution (L)
TVKA : 1.35  : Absorption rate constant (1/hr)
WT   : 70    : Weight (kg)
SEX  : 1     : Male = 0, Female 1
WTCL : 0.75  : Exponent weight on CL
SEXV : 0.878 : Volume female/Volume male
```

```{r,eval=FALSE}
[MAIN]
double CL = TVCL*pow(WT/70,WTCL)*exp(ECL);
double V  = TVV *pow(SEXVC,SEX)*exp(EV);
double KA = TVKA*exp(EKA);
```

```{r,eval=FALSE}
[OMEGA] @name OMGA @correlation @block
ECL : 1.23 : Random effect on CL
EV  : 0.67 0.4 : Random effect on V
EKA : 0.25 0.87 0.2 : Random effect on KA
```

```{r,eval=FALSE}
[SIGMA] @name SGMA 
PROP: 0.25 : Proportional residual error
ADD : 25   : Additive residual error
```

```{r,eval=FALSE}
[CMT]
GUT  : Dosing compartment  (mg)
CENT : Central compartment (mg)
```


```{r,eval=FALSE}
[PKMODEL] ncmt = 1, depot=TRUE
```

```{r,eval=FALSE}
[TABLE]
capture IPRED = CENT/V;
double DV = IPRED*(1+PROP) + ADD;

```{r,eval=FALSE}
[CAPTURE]
DV  : Concentration (mg/L)
ECL : Random effect on CL
CL  : Individual clearance (L/hr)
```


## Set initial conditions {#topic-init}

``` {r}
library(mrgsolve)
library(dplyr)
```
```{r,echo=FALSE}
options(mrgsolve_mread_quiet = TRUE)
knitr::opts_chunk$set(comment='.')
```



### Summary

-  `mrgsolve` keeps a base list of compartments and initial conditions that 
you can update **either** from `R` or from inside the model specification
-  When you use `$CMT`, the value in that base list is assumed to be 0 for 
every compartment
-  `mrgsolve` will by default use the values in that base list when starting 
the problem
-  When only the base list is available, every individual will get the same 
initial condition
-  You can **override** this base list by including code in `$MAIN` to set 
the initial condition
-  Most often, you do this so that the initial is calculated as a function 
of a parameter
-  For example, `$MAIN RESP_0 = KIN/KOUT;` when `KIN` and `KOUT` have some value in `$PARAM`
-  This code in `$MAIN` overwrites the value in the base list for the current `ID`
-  For typical PK/PD type models, we most frequently initialize in `$MAIN`
-  This is equivalent to what you might do in your NONMEM model
-  For larger systems models, we often just set the initial value via the base list

### Make a model only to examine `init` behavior


Note: `IFLAG` is my invention only for this demo. The demo is always responsible for setting and interpreting the value (it is not reserved in any way and `mrgsolve` does not control the value).

For this demo

-   Compartment `A` initial condition defaults to 0
-   Compartment `A` initial condition will get set to `BASE` **only** if `IFLAG  > 0`
-   Compartment `A` always stays at the initial condition

``` {r}
code <- '
$PARAM BASE=100, IFLAG = 0

$CMT A

$MAIN

if(IFLAG > 0) A_0 = BASE;

$ODE dxdt_A = 0;
'
```

``` {r inits-mread,cache=TRUE}
mod <- mcode("init",code)
```

**Check the initial condition**

``` {r}
init(mod)
```

Note:

-   We used `$CMT` in the model spec; that implies that the base initial 
condition for `A` is set to 0
-   In this chunk, the code in `$MAIN` doesn't get run because `IFLAG` is 0
-   So, if we don't update something in `$MAIN` the initial condition is as we 
set it in the base list

``` {r topic-init-2, cache=TRUE}
mod %>% mrgsim %>% plot
```


**Next, we update the base initial condition for `A` to 50**

Note:

-   The code in `$MAIN` still doesn't get run because `IFLAG` is 0

``` {r topic-init-3,cache=TRUE}
mod %>% init(A = 50) %>% mrgsim %>% plot
```

**Now, turn on `IFLAG`**

Note:

-   Now, that code in `$MAIN` gets run
-   `A_0` is set to the value of `BASE`

``` {r topic-init-4,cache=TRUE}
mod %>% param(IFLAG=1) %>% mrgsim %>% plot
```


``` {r topic-init-5,cache=TRUE}
mod %>% param(IFLAG=1, BASE=300) %>% mrgsim %>% plot
```


### Example PK/PD model with initial condition


Just to be clear, there is no need to set any sort of flag to set the initial 
condition.

``` {r}
code <- '
$PARAM AUC=0, AUC50 = 75, KIN=200, KOUT=5

$CMT RESP

$MAIN 
RESP_0 = KIN/KOUT;

$ODE

dxdt_RESP = KIN*(1-AUC/(AUC50+AUC)) - KOUT*RESP;

'
```

``` {r topic-init-6,cache=TRUE}
mod <- mcode("init2", code)
```

The initial condition is set to 40 per the values of `KIN` and `KOUT`

``` {r topic-init-7, cache=TRUE}
mod %>% mrgsim %>% plot
```

Even when we change `RESP_0` in `R`, the calculation in `$MAIN` gets the final say

``` {r topic-init-8, cache=TRUE}
mod %>% init(RESP=1E9) %>% mrgsim
```


### Remember: calling `init` will let you check to see what is going on

-  It's a good idea to get in the habit of doing this when things aren't clear
-  `init` first takes the base initial condition list, then calls `$MAIN` and 
does any calculation you have in there; so the result is the calculated initials

``` {r, topic-init-9}
init(mod)
```

``` {r, topic init 10}
mod %>% param(KIN=100) %>% init
```



### Set initial conditions via `idata`

Go back to house model

``` {r, init-by-idata-house}
mod <- mrgsolve:::house()
```

``` {r,init-by-idata-init}
init(mod)
```

Notes

-  In `idata` (only), include a column with `CMT_0` (like you'd do in `$MAIN`).
-  When each ID is simulated, the `idata` value will override the base initial 
list for that subject.
-  But note that if `CMT_0` is set in `$MAIN`, that will override the `idata` 
update.

``` {r,init-by-idata-expand.idata}
idata <- expand.idata(CENT_0 = seq(0,25,1))
```

``` {r,init-by-idata-expand.idata-2}
idata %>% head
```


``` {r topic-init-10,cache=TRUE}
out <- 
  mod %>% 
  idata_set(idata) %>% 
  mrgsim(end=40)
```

``` {r topic-init-11,cache=TRUE}
plot(out, CENT~.)
```


## Updating parameters {#topic-parameter-update}

The parameter list was introduced in section \@ref(component-param) and the 
`$PARAM` code block was shown in \@ref(block-param). Once a model is compiled, 
the names and number of parameters in a model is fixed. However, the values of parameters can be changed: parameters may be updated either by the user 
(in `R`) or by `mrgsolve` (in the `C++` simulation engine, as the simulation proceeds). 

* To update in `R`, use the `param()` function (see examples below)
* To have `mrgsolve` update the parameters, attach columns to your data set 
(either `data_set` or `idata_set`) with the same name as items in the parameter 
list

Both of these methods are discussed and illustrated in the following sections.

### Parameter update hierarchy

As we noted above, new parameter values can come from three potential sources:

1. Modification of the (base) parameter list
1. A column in an `idata_set` that has the same name as a model parameter
1. A column in a `data_set` that has the same name as a model parameter

These sources for new parameter values are discussed below.  We note here that
the sources listed above are listed in the order of the parameter update
*hierarchy*.  So, the base parameter list provides the value by default.  A 
parameter value coming from an `idata_set` will override the value in the 
base list.  And a parameter value coming from a `data_set` will override
the value coming from the base list or an `idata_set` (in case a parameter 
is listed in both the `idata_set` and the `data_set`).  In other words,
the hierarchy is:

1. base parameter list is the default
1. the `idata_set` overrides the base list
1. the `data_set` overrides the `idata_set` and the base list

The parameter update hierarchy is discussed in the following sections.

__Base parameter set__

* Every model has a base set of "parameters"
* These are named and set in `$PARAM`
* Parameters can only get into the parameter list in `$PARAM` (or `$THETA`)
* No changing the names or numbers of parameters once the model is compiled
* But, several ways to change the values

```{r,message=FALSE}
code <- '
$VCMT KYLE
$PARAM CL = 1.1, VC=23.1, KA=1.7, KM=10
$CAPTURE CL VC KA KM
'
mod <- mcode("tmp", code, warn=FALSE)
```

```{r}
param(mod)
```

__The base parameter set is the default__

The base parameter set allows you to run the model without entering any other 
data; there are some default values in place.  

__The parameters in the base list can be changed or updated in `R`__

Use the `param()` function to both set and get:

```{r}
mod %<>% param(CL=2.1)
```

```{r}
param(mod)
```
But whatever you've done in `R`, there is a base set (with values) to use. 
See section \@ref(topic-parameter-update-base) for a more detailed discussion 
of using `param()` to updated the base list.

__Parameters can also be updated during the simulation run__

Parameters can be updated by putting columns in `idata` set or `data_set` that 
have the same name as one of the parameters in the parameter list.  But there 
is no changing values in the base parameter set once the simulation starts.  
That is, the following model specification will not compile:

```{c,eval=FALSE}
$PARAM CL = 2

$MAIN CL = 3; // ERROR
```

You cannot over-write the value of a parameter in the model specification.  
Let `mrgsolve` do the updating.

`mrgsolve` always reverts to the base parameter set when starting work on a 
new individual.

__Parameters updated from `idata_set`__

When `mrgsolve` finds parameters in `idata`, it will update the base parameter 
list with those parameters prior to starting that individual.

```{r}
data(exidata)
head(exidata)
```

Notice that there are several columns in `exidata` that match up with the 
names in the parameter list
```{r}
names(exidata)

names(param(mod))
```

The matching names tell `mrgsolve` to update, assigning each individual 
their individual parameter.


```{r topic-pars-4,cache=TRUE}
out <- 
  mod %>%
  idata_set(exidata) %>%
  mrgsim(end=-1 , add=c(0,2))
```

```{r}
out
```

__Parameters updated from `data_set`__

Like an `idata` set, we can put parameters on a `data` set

```{r}
data <- expand.ev(amt=0, CL=c(1,2,3), VC=30)
```

```{r topic-pars-5, cache=TRUE}
out <- 
  mod %>%
  data_set(data) %>% 
  obsonly %>%
  mrgsim(end=-1, add=c(0,2))
```
```{r}
out
```

This is how we do time-varying parameters:
```{r}
data <- 
  data_frame(CL=seq(1,5)) %>%
  mutate(evid=0,ID=1,cmt=1,time=CL-1,amt=0)
```

```{r}
mod %>%
  data_set(data) %>%
  mrgsim(end=-1)
```

For more information on time-varying covariates 
(parameters), see sections \@ref(topics-time-varying) and \@ref(section-sequence).


__Parameters are carried back when first record isn't at `time == 0`__

What about this?

```{r}
data <- expand.ev(amt=100,time=24,CL=5,VC=32)
data
```

The first `data` record happens at `time==24`

```{r}
mod %>% 
  data_set(data) %>%
  mrgsim(end=-1, add=c(0,2))
```

Since the data set doesn't start until `time==5`, we might think that `CL` doesn't change from the base parameter set until then.  

But by default, `mrgsolve` carries those parameter values back to the start of the simulation.  This is by design ... by far the more useful configuration.

If you wanted the base parameter set in play until that first data set record, do this:

```{r}
mod %>%
  data_set(data) %>%
  mrgsim(end=-1,add=c(0,2), filbak=FALSE)
```

Will this work? 

```{r}

idata <- do.call("expand.idata", as.list(param(mod)))

idata

```
Here, we'll pass in __both__ `data_set` and `idata_set` and they have conflicting values for the parameters.

```{r}

mod %>%
  data_set(data) %>%
  idata_set(idata) %>%
  mrgsim(end=-1,add=c(0,2))
```

The data set always gets the last word.


### Updating the base parameter list {#topic-parameter-update-base}

From the previous section
```{r}
param(mod)
```

__Update with `name-value` pairs__

We can call `param()` to update the model object, directly naming the parameter to update and the new value to take

```{r}
mod %>% param(CL = 777, KM = 999) %>% param
```

The parameter list can also be updated by scanning the names in a list

```{r}
what <- list(CL = 555, VC = 888, KYLE = 123, MN = 100)

mod %>% param(what) %>% param
```

`mrgsolve` looks at the names to drive the update.  `KYLE` (a compartment name) and `MN` (not in the model anywhere) are ignored.

Alternatively, we can pick a row from a data frame to provide the input for the update
```{r}
d <- data_frame(CL=c(9,10), VC=c(11,12), KTB=c(13,14))

mod %>% param(d[2,]) %>% param

```

Here the second row in the data frame drives the update.  Other names are ignored.  

A warning will be issued if an update is attempted, but no matching names are found
```{r,eval=FALSE}
mod %>% param(ZIP = 1, CODE = 2) %>% param
```

```
Warning message:
Found nothing to update: param 
```



## Time grid objects {#topic-tgrid}

__Simulation times in `mrgsolve`__

```{r}
mod <- mrgsolve:::house() %>% Req(CP) %>% ev(amt=1000,ii=24, addl=1000) 
```


`mrgsolve` keeps track of a simulation `start` and `end` time and a 
fixed size step between `start` and `end` (called `delta`).  `mrgsolve`
also keeps an arbitrary vector of simulation times called `add`. 

```{r}
mod %>%
  mrgsim(end=4,delta=2,add=c(7,9,50)) %>%
  as.data.frame
```


__`tgrid` objects__

The `tgrid` object abstracts this setup and allows us to make 
complicated sampling designs from elementary building blocks.

__Make a day 1 sampling with intensive sampling around the peak and sparser otherwise__
```{r}
peak1 <- tgrid(1,4,0.1)
sparse1 <- tgrid(0,24,4)
```

__Use the `c` operator to combine simpler designs into more complicated designs__
```{r}
day1 <- c(peak1,sparse1)
```

Check this by calling `stime`
```{r}
stime(day1)
```

Pass this object in to `mrgsim` as `tgrid`.  It will override the default `start/end/delta/add` sequence.
```{r topic-tgrid-1, cache=TRUE}
mod %>% 
  mrgsim(tgrid=day1) %>%
  plot(type='b')
```


__Now, look at both day 1 and day 10__:

Adding a number to a `tgrid` object will offset those times by that amount.  

```{r topic-tgrid-2, cache=TRUE}
des <- c(day1, day1+10*24)

mod %>% 
  mrgsim(tgrid=des) %>%
  plot(type='b')


```

Pick up day 5 as well

```{r topic-tgrid-3, cache=TRUE}
des <- c(des, day1+5*24)

mod %>% 
  mrgsim(tgrid=des) %>%
  plot(type='b')
```



## Individualized sampling designs {#topic-designs}

Here is a PopPK model and a full `data_set`.

```{r}
mod <- mrgsolve:::house()

data(exTheoph)

df <- exTheoph

head(df)
```

```{r topic-tgrid-5,cache=TRUE}
mod %>% 
  Req(CP) %>%
  carry.out(a.u.g) %>%
  data_set(df) %>%
  obsaug %>%
  mrgsim 
```

Now, define two time grid objects: `des1` runs from 0 to 24 and `des2` runs from 0 to 96, both every hour.
```{r}

des1 <- tgrid(0,24,1)
des2 <- tgrid(0,96,1)

range(stime(des1))
range(stime(des2))
```

Now, derive an `idata_set` after adding a grouping column (`GRP`) that splits the data set into two groups
```{r}
df %<>% mutate(GRP = as.integer(ID > 5))

id <- df %>% distinct(ID,GRP) 

id
```

Now, we have two groups in `GRP` in `idata_set` and we have two `tgrid` objects.  

* Pass in both the `idata_set` and the `data_set`
* Call `design`
* Identify `GRP` as `descol`; the column __must__ be in `idata_set`
* Pass in a list of designs; it __must__ be at least two because there are two levels in `GRP`


When we simulate, the individuals in `GRP 1` will get `des1` and those in `GRP 2` will get `des2`

```{r topic-design-1,cache=TRUE}
out <- 
  mod %>% 
  Req(CP) %>%
  carry.out(a.u.g,GRP) %>%
  idata_set(id) %>%
  data_set(df) %>%
  design(descol="GRP", deslist=list(des1,des2)) %>%
  obsaug %>%
  mrgsim 

plot(out, CP~time|GRP)
```


## Some helpful `C++`

```{c,eval=FALSE}

// If
if(a == 2) b = 2;

// If else
if(b <= 2) {
c=3;
} else {
c=4;
}

// x <- ifelse(c == 4, 8, 10)
double x = c == 4 ? 8 : 10;

double d = pow(base,exponent);
double e = exp(3);
double f = fabs(-4);
double g = sqrt(5);
double h = log(6);
double i = log10(7);
double j = floor(4.2);
double k = ceil(4.2);
```

## Resimulate `ETA` and `EPS`


__Call `simeps()` to resimulate `ETA`__

- No `$PLUGIN` is required
- `simeta()` takes no arguments

For example, we can simulate individual-level covariates
within a certain range:

```{r topic-simeta-1, cache=TRUE}
code <- '
$PARAM TVCL = 1, TVWT = 70

$MAIN 
capture WT = TVWT*exp(EWT);

int i = 0;

while((WT < 60) || (WT > 80)) {
  if(++i > 100) break;
  simeta();
  WT = TVWT*exp(EWT);
}

$OMEGA @labels EWT
4

$CAPTURE EWT WT
'

mod <- mcode("simeta", code)

out <- mod %>% mrgsim(nid=100, end=-1)

sum <- summary(out)

sum
```


__Call `simeps()` to resimulate `EPS`__

- No `$PLUGIN` is required
- `simeps()` takes no arguments

For example, we can resimulate untill all 
concentrations are greater than zero:

```{r topic-simeps-1,cache=TRUE}
code <- '
$PARAM CL = 1, V = 20,

$CMT CENT

$SIGMA 50

$PKMODEL ncmt=1

$TABLE
capture CP = CENT/V + EPS(1);

int i = 0;

while(CP < 0 && i < 100) {
  simeps();
  CP = CENT/V + EPS(1);
  ++i;
}

'

mod <- mcode("simeps", code)

out <- mod %>% ev(amt=100) %>% mrgsim(end=48)
sum <- summary(out)

sum
```


__A safety check is recommended__
Note that in both examples, we implement a safety check: an integer 
counter is incremented every time we resimulated.  The resimulation 
process stops if we don't reach the desired condition within 
100 replicates.  You might also consider issuing a message
or a flag in the simulated data if you are not able to reach the desired condition.

## Updating `$OMEGA` and `$SIGMA` \  {#topic-matrix-update}

Like the values of parameters in the parameter list, we may want to update the 
values in `$OMEGA` and `$SIGMA` matrices.  We can do so without re-compiling the 
model.

### Matrix helper functions

`mrgsolve` keeps `$OMEGA` and `$SIGMA` in block matrices (regardless of whether the off-diagonal elements are zeros or not).  Recall that in the model specification file
we can enter data for `$OMEGA` and `$SIGMA` as the lower triangle of the matrix (see section \@ref(block-omega)).  In `R`, we need to provide a matrix (as an `R` object).  `mrgsolve` provides some convenience functions to help ... allowing the user to enter lower diagonals instead of the full matrix. 


`dmat()` for diagonal matrix
```{r}
dmat(1,2,3)
```

`bmat()` for block matrix
```{r}
dmat(1,2,3)
```

`cmat()` for a block matrix where the diagonal elements are variances and the off-diagonals are taken to be correlations, not covariances
```{r}
cmat(0.1, 0.87,0.3)
```
`mrgsolve` will convert the correlations to covariances. 


`mrgsolve` also provides `as_bmat()` and `as_dmat()` for converting other `R` objects to matrices or lists of matrices.  

Consider this list with named elements holding the data for a matrix:
```{r}
m <- list(OMEGA1.1 = 0.9, OMEGA2.1 = 0.3, OMEGA2.2 = 0.4)
```

These data could form either a 3x3 diagonal matrix or a 2x2 block matrix.  But the names suggest a 2x2 form.  `as_bmat()` can make the matrix like this

```{r}
as_bmat(m, "OMEGA")
```

The second argument is a regular expression that `mrgsolve` uses to find elements in the list to use for building the matrix.

Frequently, we have estimates in a data frame like this
```{r}
data(exBoot)
head(exBoot)
```

We can use `as_bmat()` with this data frame to extract the `$OMEGA` matrices
```{r}
omegas <- as_bmat(exBoot, "OMEGA")
length(omegas)
dim(exBoot)

omegas[[6]]

omegas[[16]]
```

The result of calling `as_bmat` or `as_dmat` is a list of matrices, one for each row in the data frame.

Note in this example, we could have called
```{r,eval=FALSE}
sigmas <- as_bmat(exBoot,"SIGMA") 
```
to grab the `$SIGMA` matrices.


For help on these helper functions, see `?dmat`, `?bmat`, `?cmat`, `?as_bmat`, 
`?as_dmat` in the `R` help system after loading `mrgsolve`.

## Fill a matrix with zeros
Sometimes we write a population model that includes random effects, but we would
like to simulate from that same model without the random effects implemented.  
For example, we want to simulate some typical PK profiles from a population
PK model that includes IIV on some parameters and / or RUV 
on the simulated outputs. 

To do this, pass the model through the `zero_re()` function.  By default, 
this will convert all `$OMEGA` and `$SIGMA` matrix elements to zeros. 
See the R help file (`?zero_re`) to see some options for selectively 
zeroing out only one or the other. 

For example we have this population PK model

```{r}
mod <- mread("popex", modlib(), compile = FALSE)

omat(mod)
```

We can turn that matrix to all zeros with

```{r}
mod %>% zero_re() %>% omat
```

And when we simulate right after that, all `ETA(n)` will be zero as well 
and you'll get your fixed-effects simulation (the following is for example
only and is not evaluated)

```{r,eval = FALSE}
mod %>% 
  zero_re() %>% 
  ev(amt = 100) %>% 
  mrgsim() %>%
  plot()
```


### Example: unnamed matrix
Here is a model with only a 3x3 `$OMEGA` matrix
```{r}
code <- '
$OMEGA
1 2 3
'

mod <- mcode("matrix", code, compile=FALSE)
```

Let's check the values in the matrix using `omat()`
```{r}
mod %>% omat
```

We also use `omat()` to update the values in the matrix 

```{r}
mod %>% omat(dmat(4,5,6)) %>% omat
```

To update `$OMEGA`, we must provide a matrix of the same dimension, in this case 3x3. An error is generated if we provide a matrix with the wrong dimension.

```{r}
ans <- try(mod %>% omat(dmat(11,23)))

ans
```

### Example: named matrices
When there are multiple `$OMEGA` matrices, it can be helpful to assign them names.  Here, there are two matrices: one for interindividual variability (`IIV`) and one for interoccasion variability (`IOV`). 

```{r}
code <- '
$OMEGA @name IIV
1 2 3
$OMEGA @name IOV
4 5
'
mod <- mcode("iov", code, compile=FALSE)

revar(mod)
```

Now, we can update either `IIV` or `IOV` (or both) by name

```{r}
mod %>% 
  omat(IOV = dmat(11,12), IIV = dmat(13,14,15)) %>% 
  omat
```

Again, an error is generated if we try to assign a 3x3 matrix to the `IOV` position
```{r}
ans <- try(mod %>% omat(IIV=dmat(1,2)))
ans
```



### Example: unnamed matrices
If we do write the model with unnamed matrices, we can still update them
```{r}
code <- '
$OMEGA
1 2 3

$OMEGA 
4 5
'
mod <- mcode("multi", code, compile=FALSE)
```

In this case, the only way to update is to pass in a __list__ of matrices, where (in this example) the first matrix is 3x3 and the second is 2x2

```{r}
mod %>% omat(list(dmat(5,6,7),dmat(8,9))) %>% omat
```


## Time varying covariates {#topics-time-varying}

A note in a previous section showed how to 
implement time-varying covariates or other
time-varying parameters by including those
parameters as column in the data set. 

By default, `mrgsolve` performs next observation
carried backward (`nocb`) when processing 
time-varying covariates.  That is, when the 
system advances from `TIME1` to `TIME2`, 
and the advance is a function of a covariate
found in the data set, the system advances 
using the covariate value `COV2` rather than
the covariate `COV1`.  

The user can change the behavior to last observation
carried forward (`locf`), so that the system 
uses the value of `COV1` to advance from `TIME1` to 
`TIME2`.  To use `locf` advance, set `nocb` to `FALSE` when calling `mrgsim`.  For example, 

```{r,eval = FALSE}
mod %>% mrgsim(nocb = FALSE)
```


Note that time-varying covariates are not 
possible when using `qsim` simulation. 

There is additional information about the sequence
of events that takes place during system advance 
in section \@ref(section-sequence).


















