# Simulated output {#simulated-output}

## Output types

When `mrgsim()` is used to simulate from a model, it by default returns an object with class 
`mrgsims`.  This is an S4 object containing a data.frame of simulated output and a 
handful of other pieces of data related to the simulation run that can be coerced to other
types (like `data.frame` or `tibble`). 

For simulations with large outputs or extremely brief simulations where efficiency is 
important, users can request the output be returned as a data frame. This is most efficient
when the features provided by the `mrgsims` object are not needed. To do this, pass
the `output` argument to `mrgsim()`

```{r, eval = FALSE}
out <- mrgsim(mod, ..., output = "df")
```
or use `mrgsim_df()`

```{r, eval = FALSE}
out <- mrgsim_df(mod, ....)
```

## Methods for mrgsim output

`mrgsolve` provides several methods for working with `mrgsims` objects or 
coercing the simulation matrix into other `R` objects. Note the discussion in the 
following subsections all refer to working with `mrgsims` objects, not data.frame output.

### Coercion methods

* `as_tibble()`: convert to `tibble`
* `as.data.frame()`: convert to `data.frame`
* `as.matrix()`: convert to `matrix`

### Query methods
* `head()`: shows the first `n = 5` rows
* `tail()`: shows the last `n = 5` rows
* `names()`: shows the column names
* `dim()`: shows the number of rows and columns
* `summary()`: shows a numeric summary of all columns
* `$`: extracts a column

### Graphical methods

There is a `plot()` methods for simulated output that is aware of independent and dependent
variables from the simulation. If `out` is the simulated output (an `mrgsims` object)

```{r, eval = FALSE}
plot(out)
```
Plot with a formula; the following example selects only the `CP` and `RESPONSE` outputs
and plots them versus `time`

```{r, eval = FALSE}
plot(out, CP + RESPONSE ~ time)
```

To select a large number of responses to plot, pass a character vector or comma-separated
character data containing output columns to plot

```{r, eval = FALSE}
plot(out, "CP, RESPONSE, WT, DOSE")
```


### Methods for `dplyr` verbs
`mrgsolve` provides several S3 methods to make it possible to include `dplyr` verbs in 
your simulation pipeline. 

For example

```{r, eval = FALSE}
library(dplyr)
library(mrgsolve)

mod <- mrgsolve:::house()

mod %>% 
  ev(amt=100) %>%
  mrgsim() %>% 
  filter(time >= 10)
```

Here, `mrgsim()` returns an `mrgsims` object.  When `dplyr` is also loaded, this object 
can be piped directly to `dplyr::filter()` or `dplyr::mutate()` etc.

It is important to note that when `mrgsims` output is piped to `dplyr` functionality, 
it is coerced to `tibble` (`data.frame`) and there is no way to get the data back to 
`mrgsims` object. Most of the time, this is desirable and there is no need to explicitly
coerce to `tibble()` when calling `dplyr` verbs on simulated output. 


Other `dplyr` functions that can be used with `mrgsims` objects

* `group_by()`
* `mutate()`
* `filter()`
* `summarise()`
* `select()`
* `slice()`
* `pull()`
* `distinct()`
* `slice()`

### Modify methods

You can modify the underlying data in the `mrgsims` object and keep it as an `mrgsims` object.

- `filter_sims()`: calls `dplyr::filter()` to pick rows to keep or discard
- `select_sims()`: calls `dplyr::select()`; note that `ID` and `time` columns are always retained
- `mutate_sims()`: calls `dplyr::mutate()` to add or modify columns

## Controlling output scope

### Background

Limiting the volume of simulated data can have a major impact on simulation efficiency, 
memory footprint, and ease (or lack of ease) in reviewing and dealing with the output.
For any large simulation or any simulation from a large model, the user should consider
selecting what gets returned when the simulation is performed.

By default, `mrgsim()` returns a data.frame with the following

1. `ID`: regardless of whether you simulated a population or not
1. `time` / `TIME`: the independent variable
1. Simulated values for all model compartments
1. Simulated values for derived outputs listed in `$CAPTURE`

You will always get `ID` and `time` and the compartments and any captured items must
be written into the model file. This defines the list of data items that 
**could (possibly)** get returned under items 3 and 4 above. Again: this must be 
written into the model file and is locked at the time the model is compiled. 

However, `mrgsolve` allows the user to pick what is actually returned at run time. 
Because this is done at run time, different runs can return different data items. 
And (importantly) `mrgsim()` only allocates space in the output for data items 
that are requested. So, opting **out** of unneeded outputs will decrease memory 
consumption and increase efficiency.

### Implementation

The mrgsolve model object tracks compartments and captures that are currently 
being requested. This can be queried using `outvars()`

```{r}
mod <- house()

outvars(mod)
```

Items are listed under `cmt` and `capture`. The user can update the model object
with the names of columns that are being requested by passing `outvars` to `update()`

```{r}
mod <- update(mod, outvars = "CP, RESP")

outvars(mod)
```

This will exclude anything that isn't named in the update. The `outvars` list can 
be reset by passing `(all)`

```{r}
mod <- update(mod, outvars = "(all)")
```

Remember that `...` passed to `mrgsim()` are also passed to `update()` so it is possible
to select outputs right in your `mrgsim()` call

```{r, eval = FALSE}
out <- mrgsim(mod, outvars = "CP, RESP")
```

### Copy items from data to simulated output

Users can also use `carry_out` and `recover` to copy items from the input data into the
output. This is covered in a different chapter. 



